Supplementary Information: Tropical tree drought sensitivity is jointly shaped by drought characteristics, species adaptations, and individual microenvironments

### 2025-04-03 doc/manuscript_hkk_drought_sensitivity.Rmd
current figure numbers from SI into table and text #22

Drought years were associated with lower growth across the plot. 
The drought year 2015 had the lowest median annual increment `r round(median(tree.time$inc_annual[tree.time$yr==2015], na.rm=T),2)` cm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2015], na.rm=T)-median(tree.time$inc_annual, na.rm=T))/median(tree.time$inc_annual, na.rm=T)),2)`% lower than the median across all years. 
In 2010, the median annual increment was `r round(median(tree.time$inc_annual[tree.time$yr==2010], na.rm=T),2)` cm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2010], na.rm=T)-median(tree.time$inc_annual, na.rm=T))/median(tree.time$inc_annual, na.rm=T)),2)`% lower than the median across all years. 
While there were more species differences in median responses to the 2010 drought, all species had lower increment than a typical year in 2015. 
Besides decrease in magnitude of growth, the proportion of trees with negative growth was also higher in the drought years (Fig S--). 
Across the years, on average, `r round(100*mean(tapply(tree.time$inc_annual, tree.time$yr, function(x) sum(x < 0) / length(x))),2)`% of the trees had negative growth, with a minimum of `r round(100*min(tapply(tree.time$inc_annual, tree.time$yr, function(x) sum(x < 0) / length(x))),2)`% in `r names(sort(tapply(tree.time$inc_annual, tree.time$yr, function(x) sum(x < 0) / length(x))))[1]`. The drought year 2015 had the maximum proportion of trees with negative growth in the timeseries, `r round(100*sum(tree.time$inc_annual[tree.time$yr==2015]<0)/nrow(tree.time[tree.time$yr==2015,]),2)`%.

@@

Deciduousness was associated with lower negative sensitivities in 2015, but not in 2010 (Fig 3a). 
Species median sensitivities in 2010 and 2015 were not correlated with each other (Pearson's correlation coefficient r = `r round(sens.sp.cor[4]$estimate, 2)`, p = `r round(sens.sp.cor[3]$p.value,2)`, Fig S--) but on average had lower mean and median sensitivities in 2015 than in 2010 (Fig S--), in alignment with median sensitivities across all trees. 
In 2015, predicted sensitivities from intercept-only models of sensitivity with a species random effect on the intercept showed a significant negative association with deciduousness value (r = `r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2015], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2015], 2)`). 
On the other hand, in 2010, predicted sensitivities had no significant association with deciduousness (r = `r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2010], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2010], 2)`). 
Moreover, models of all trees with TWI and deciduousness as predictors showed that the conditional effect of deciduousness on sensitivity (controlling for topographic wetness) was significant and negative in 2015 (`r round(isocline_coefs_dec$median[isocline_coefs_dec$yr==2015],2)`; 90% CI `r round(isocline_coefs_dec$lwr[isocline_coefs_dec$yr==2015],2)`, `r round(isocline_coefs_dec$upr[isocline_coefs_dec$yr==2015],2)`) but not in 2010 (`r round(isocline_coefs_dec$median[isocline_coefs_dec$yr==2010],2)`, 90% CI `r round(isocline_coefs_dec$lwr[isocline_coefs_dec$yr==2010],2)`, `r round(isocline_coefs_dec$upr[isocline_coefs_dec$yr==2010],2)`).

@@

The effect of wetness on sensitivity was different in 2010 and 2015, with 2015 effects interacting with deciduousness. 
Across all trees, the model with TWI and deciduousness showed that TWI had no effect on drought sensitivity in 2010 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2010],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2010],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2010],2)`) but a positive effect in 2015 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2015],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2015],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2015],2)`). 
The interaction between TWI and deciduousness was not significant in 2010 (median effect =`r round(isocline_coefs_inter$median[isocline_coefs_inter$yr==2010],2)`, 90% CI `r round(isocline_coefs_inter$lwr[isocline_coefs_inter$yr==2010],2)`, `r round(isocline_coefs_inter$upr[isocline_coefs_inter$yr==2010],2)`) but was significantly negative in 2015 (median effect =`r round(isocline_coefs_inter$median[isocline_coefs_inter$yr==2015],2)`, 90% CI `r round(isocline_coefs_inter$lwr[isocline_coefs_inter$yr==2015],2)`, `r round(isocline_coefs_inter$upr[isocline_coefs_inter$yr==2015],2)`).
Combined, predicted sensitivities of deciduous trees was more negative than evergreen trees in 2015, with this effect stronger in wetter sites (Fig 3b).

@@

<!----Contributions beyond---->
Despite consistent observations of growth declines at the plot level across drought events, we show that this scale obscures a variety of ecological responses at the species and tree level that are heterogeneous across drought events. 
Our finding that a wet season drought had bigger impact calls into question the generality of the findings of @zuidema_tropical_2022, which found that in species with tree-ring chronologies, dry season drought was most impactful. 
(Care must be taken in interpretation, however, as the 2015-2016 wet season ENSO drought was stronger (Fig 1)). 
<!-- Given the limitations of tree ring methods and tape measurements in measuring accurately measuring growth differences across a range of species in the tropics, methods so far could have led to capturing only the most dramatic droughts that affected (dry-associated) tree ring species and fast-growing individuals.  -->
*Afzelia xylocarpa*, one of the few annual-ring-forming species in HKK is noteworthy with the most distinct response across the two events; it had the highest (positive) sensitivity in the dry season drought, but a large negative sensitivity in the wet season drought (Fig S--). 
More generally, past analyses of tree ring and census estimates of drought sensitivity in seasonally dry tropical forests may have been limited towards fast growing (measurable using tape) or few dry-associated species (that form annual tree rings) - and therefore capturing only the more dramatic effects. 
In contrast, our analyses reveal a range of responses, including the contrasting influence of deciduousness in the two different droughts, showing that ecological and evolutionary strategies may result in heterogeneous consequences. 
Distinct responses of species and individuals to drought characteristics suggests potential spatiotemporal insurance effects among tropical tree response to drought in alignment with recent global analyses (@liu_drought_2022).

>>>>

Drought years were associated with lower growth across the plot. 
The drought year 2015 had the lowest median annual increment `r round(median(tree.time$inc_annual[tree.time$yr==2015], na.rm=T),2)` cm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2015], na.rm=T)-median(tree.time$inc_annual, na.rm=T))/median(tree.time$inc_annual, na.rm=T)),2)`% lower than the median across all years. 
In 2010, the median annual increment was `r round(median(tree.time$inc_annual[tree.time$yr==2010], na.rm=T),2)` cm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2010], na.rm=T)-median(tree.time$inc_annual, na.rm=T))/median(tree.time$inc_annual, na.rm=T)),2)`% lower than the median across all years. 
While there were more species differences in median responses to the 2010 drought, all species had lower increment than a typical year in 2015. 
Besides decrease in magnitude of growth, the proportion of trees with negative growth was also higher in the drought years (Fig S4). 
Across the years, on average, `r round(100*mean(tapply(tree.time$inc_annual, tree.time$yr, function(x) sum(x < 0) / length(x))),2)`% of the trees had negative growth, with a minimum of `r round(100*min(tapply(tree.time$inc_annual, tree.time$yr, function(x) sum(x < 0) / length(x))),2)`% in `r names(sort(tapply(tree.time$inc_annual, tree.time$yr, function(x) sum(x < 0) / length(x))))[1]`. The drought year 2015 had the maximum proportion of trees with negative growth in the timeseries, `r round(100*sum(tree.time$inc_annual[tree.time$yr==2015]<0)/nrow(tree.time[tree.time$yr==2015,]),2)`%.

@@

Deciduousness was associated with lower negative sensitivities in 2015, but not in 2010 (Fig 3a). 
Species median sensitivities in 2010 and 2015 were not correlated with each other (Pearson's correlation coefficient r = `r round(sens.sp.cor[4]$estimate, 2)`, p = `r round(sens.sp.cor[3]$p.value,2)`, Fig S5) but on average had lower mean and median sensitivities in 2015 than in 2010 (Fig S5), in alignment with median sensitivities across all trees. 
In 2015, predicted sensitivities from intercept-only models of sensitivity with a species random effect on the intercept showed a significant negative association with deciduousness value (r = `r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2015], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2015], 2)`). 
On the other hand, in 2010, predicted sensitivities had no significant association with deciduousness (r = `r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2010], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2010], 2)`). 
Moreover, models of all trees with TWI and deciduousness as predictors showed that the conditional effect of deciduousness on sensitivity (controlling for topographic wetness) was significant and negative in 2015 (`r round(isocline_coefs_dec$median[isocline_coefs_dec$yr==2015],2)`; 90% CI `r round(isocline_coefs_dec$lwr[isocline_coefs_dec$yr==2015],2)`, `r round(isocline_coefs_dec$upr[isocline_coefs_dec$yr==2015],2)`) but not in 2010 (`r round(isocline_coefs_dec$median[isocline_coefs_dec$yr==2010],2)`, 90% CI `r round(isocline_coefs_dec$lwr[isocline_coefs_dec$yr==2010],2)`, `r round(isocline_coefs_dec$upr[isocline_coefs_dec$yr==2010],2)`).

@@

The effect of wetness on sensitivity was different in 2010 and 2015, with 2015 effects interacting with deciduousness. 
Individual tree responses to drought were not correlated with each other (Fig S6). 
Across all trees, the model with TWI and deciduousness showed that TWI had no effect on drought sensitivity in 2010 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2010],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2010],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2010],2)`) but a positive effect in 2015 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2015],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2015],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2015],2)`). 
The interaction between TWI and deciduousness was not significant in 2010 (median effect =`r round(isocline_coefs_inter$median[isocline_coefs_inter$yr==2010],2)`, 90% CI `r round(isocline_coefs_inter$lwr[isocline_coefs_inter$yr==2010],2)`, `r round(isocline_coefs_inter$upr[isocline_coefs_inter$yr==2010],2)`) but was significantly negative in 2015 (median effect =`r round(isocline_coefs_inter$median[isocline_coefs_inter$yr==2015],2)`, 90% CI `r round(isocline_coefs_inter$lwr[isocline_coefs_inter$yr==2015],2)`, `r round(isocline_coefs_inter$upr[isocline_coefs_inter$yr==2015],2)`).
Combined, predicted sensitivities of deciduous trees was more negative than evergreen trees in 2015, with this effect stronger in wetter sites (Fig 3b).

@@

<!----Contributions beyond---->
Despite consistent observations of growth declines at the plot level across drought events, we show that this scale obscures a variety of ecological responses at the species and tree level that are heterogeneous across drought events. 
Our finding that a wet season drought had bigger impact calls into question the generality of the findings of @zuidema_tropical_2022, which found that in species with tree-ring chronologies, dry season drought was most impactful. 
(Care must be taken in interpretation, however, as the 2015-2016 wet season ENSO drought was stronger (Fig 1)). 
<!-- Given the limitations of tree ring methods and tape measurements in measuring accurately measuring growth differences across a range of species in the tropics, methods so far could have led to capturing only the most dramatic droughts that affected (dry-associated) tree ring species and fast-growing individuals.  -->
*Afzelia xylocarpa*, one of the few annual-ring-forming species in HKK is noteworthy with the most distinct response across the two events; it had the highest (positive) sensitivity in the dry season drought, but a large negative sensitivity in the wet season drought (Fig S5). 
More generally, past analyses of tree ring and census estimates of drought sensitivity in seasonally dry tropical forests may have been limited towards fast growing (measurable using tape) or few dry-associated species (that form annual tree rings) - and therefore capturing only the more dramatic effects. 
In contrast, our analyses reveal a range of responses, including the contrasting influence of deciduousness in the two different droughts, showing that ecological and evolutionary strategies may result in heterogeneous consequences. 
Distinct responses of species and individuals to drought characteristics suggests potential spatiotemporal insurance effects among tropical tree response to drought in alignment with recent global analyses (@liu_drought_2022).
